Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students

### 2023-06-12 phd-cs-us.tex
updates

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science \\Ph.D. Admission in US Universities\\{\large A guide for Vietnamese and International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen}\\
\href{https://cs.gmu.edu}{George Mason University}}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

@@

Prestigious fellowships typically require a clear and good research plan, so it is a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top Ph.D. students at top universities worldwide.

\begin{commentbox}
At GMU, Ph.D. applicants are automatically eligible for a Presidential Fellowship.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you are not depending on any prof. or TA duties).  PhD admission committee members nominate applicants for this fellowship and the committee will vote and give the fellowship to the top 2.
\end{commentbox}

@@

  \begin{table}[h!]
    \centering
    \small
    \caption{Top 50 CS PhD programs in the U.S. (CSRankings, June 2023). \red{$^*$} indicates the university has a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS Ph.D. students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Univ. of California - Irvine \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 &  Duke University \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & MIT & 29 & Univ. of California - Riverside\\
    5 & Georgia Institute of Technology         & 30 & Northwestern University\\
    6 & Stanford University& 31 & Pennsylvania State University  \\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\  
    8 & University of Washington      &33 &  Harvard University \\
    9 &  Univ. of California - Berkeley  &34&  Univ. of California - Santa Cruz \\
    10 & Cornell University  & 35 &  Yale University \\
    11 & University of Maryland - College Park &  36& Brown University \\ 
    12 & Northeastern University\red{$^*$} &37&  Ohio State University\\
    13 & University of Wisconsin - Madison\red{$^*$}  &38& Texas A\&M University\red{$^*$} \\
    14 & Columbia University   &39 & Boston University  \\
    15 &   Purdue University  &40& North Carolina State University\\\  
    16 & University of Texas at Austin   &41 & University of Utah \\
    17 & University of Pennsylvania\red{$^*$} &42 & University at Buffalo\red{$^*$}\\
    18 & Princeton University  & 43& Rice University\\
    19 & University of Massachusetts-Amherst\red{$^*$} & 44&  University of Colorado-Boulder \\
    20 &  New York University  &45& University of Illinois at Chicago  \\
    21 & Univ. of California - Los Angeles &46& Virginia Tech\red{$^*$}  \\
    22 & University of Southern California &47&  Arizona State University\red{$^*$} \\
    23 & Stony Brook University\red{$^*$} &48&University of Minnesota \\
    24 & University of Chicago &49& University of Virginia \\
    25 & Univ. of California - Santa Barbara &50& University of North Carolina\red{$^*$} \\
    \bottomrule
    \end{tabular}
\end{table}

@@

# ** Others  
# - There are also various kind of one-time thing and small awards from industry or organization, e.g., a $2000 award for the summer.
Unlike the B.A., where you or your parents pay many tens of thousands of dollars,
or the M.S., where you typically work as a teaching assistant and possibly continue
to pay many tens of thousands of dollars, the Ph.D. is a time where funding is not a
concern to you. At most schools, you will not pay tuition during the time that you
are getting a Ph.D.. Typically, you will also receive a living stipend – on the order
of $2000 per month, from which you will pay your living expenses. Ideally, your
only responsibility will be research. This is called doing an RAship (Research
Assistantship).
The Ph.D. is a tremendous opportunity. You get to pick an advisor in any re-
search area you like and then you get to do research in that area, receive mentoring,
think deeply on problems, publish papers, become famous, while paying zero tu-
ition for 6 years and receiving a salary. Your advisor is paying for this opportunity
by writing grant proposals to companies and to the government to ask for fund-
ing. A single graduate student can cost an advisor upwards of 80K per year (given
the cost of tuition, the stipend, the overhead tax charged by the school, cost of
equipment and physical space, etc.).
Important note 1: At most schools, you can only do an RAship if you have
an advisor who has funding for you. Since some advisors don’t apply for grants
or are in areas which aren’t well-funded, you may have to work as a teaching
assistant every semester to get your stipend. This is called a TAship (Teaching
Assistantship). When I was a graduate student, I had a few friends who were
forced to TA 13 semesters, to fund their way through school! Alternatively, you
will have to restrict your choice of advisors to those who have funding. At CMU,
every Ph.D. student is guaranteed a stipend plus tuition regardless of which advisor
she chooses to work with.
Important note 2: There are many companies and government organizations
which offer Graduate Fellowships for Ph.D. students. If you are lucky enough
to get one of these, they will cover most of your way through graduate school,
and you will never have to worry about whether your advisor has funding or not.
Details about graduate fellowships will be discussed in Section 4.

>>>>

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen}\\
\href{https://cs.gmu.edu}{George Mason University}}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

@@

Prestigious fellowships typically require a clear and good research plan, so it is a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

\begin{commentbox}
At GMU, PhD applicants are automatically eligible for a Presidential Fellowship.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you are not depending on any prof. or TA duties).  PhD admission committee members nominate applicants for this fellowship and the committee will vote and give the fellowship to the top 2.
\end{commentbox}

@@

  \begin{table}[h!]
    \centering
    \small
    \caption{Top 50 CS PhD programs in the U.S. (CSRankings, June 2023). \red{$^*$} indicates the university has a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Univ. of California - Irvine \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 &  Duke University \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & MIT & 29 & Univ. of California - Riverside\\
    5 & Georgia Institute of Technology         & 30 & Northwestern University\\
    6 & Stanford University& 31 & Pennsylvania State University  \\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\  
    8 & University of Washington      &33 &  Harvard University \\
    9 &  Univ. of California - Berkeley  &34&  Univ. of California - Santa Cruz \\
    10 & Cornell University  & 35 &  Yale University \\
    11 & University of Maryland - College Park &  36& Brown University \\ 
    12 & Northeastern University\red{$^*$} &37&  Ohio State University\\
    13 & University of Wisconsin - Madison\red{$^*$}  &38& Texas A\&M University\red{$^*$} \\
    14 & Columbia University   &39 & Boston University  \\
    15 &   Purdue University  &40& North Carolina State University\\\  
    16 & University of Texas at Austin   &41 & University of Utah \\
    17 & University of Pennsylvania\red{$^*$} &42 & University at Buffalo\red{$^*$}\\
    18 & Princeton University  & 43& Rice University\\
    19 & University of Massachusetts-Amherst\red{$^*$} & 44&  University of Colorado-Boulder \\
    20 &  New York University  &45& University of Illinois at Chicago  \\
    21 & Univ. of California - Los Angeles &46& Virginia Tech\red{$^*$}  \\
    22 & University of Southern California &47&  Arizona State University\red{$^*$} \\
    23 & Stony Brook University\red{$^*$} &48&University of Minnesota \\
    24 & University of Chicago &49& University of Virginia \\
    25 & Univ. of California - Santa Barbara &50& University of North Carolina\red{$^*$} \\
    \bottomrule
    \end{tabular}
\end{table}

@@

# ** Others  
# - There are also various kind of one-time thing and small awards from industry or organization, e.g., a $2000 award for the summer.
Unlike the B.A., where you or your parents pay many tens of thousands of dollars,
or the M.S., where you typically work as a teaching assistant and possibly continue
to pay many tens of thousands of dollars, the PhD is a time where funding is not a
concern to you. At most schools, you will not pay tuition during the time that you
are getting a PhD. Typically, you will also receive a living stipend – on the order
of $2000 per month, from which you will pay your living expenses. Ideally, your
only responsibility will be research. This is called doing an RAship (Research
Assistantship).
The PhD is a tremendous opportunity. You get to pick an advisor in any re-
search area you like and then you get to do research in that area, receive mentoring,
think deeply on problems, publish papers, become famous, while paying zero tu-
ition for 6 years and receiving a salary. Your advisor is paying for this opportunity
by writing grant proposals to companies and to the government to ask for fund-
ing. A single graduate student can cost an advisor upwards of 80K per year (given
the cost of tuition, the stipend, the overhead tax charged by the school, cost of
equipment and physical space, etc.).
Important note 1: At most schools, you can only do an RAship if you have
an advisor who has funding for you. Since some advisors don’t apply for grants
or are in areas which aren’t well-funded, you may have to work as a teaching
assistant every semester to get your stipend. This is called a TAship (Teaching
Assistantship). When I was a graduate student, I had a few friends who were
forced to TA 13 semesters, to fund their way through school! Alternatively, you
will have to restrict your choice of advisors to those who have funding. At CMU,
every PhD student is guaranteed a stipend plus tuition regardless of which advisor
she chooses to work with.
Important note 2: There are many companies and government organizations
which offer Graduate Fellowships for PhD students. If you are lucky enough
to get one of these, they will cover most of your way through graduate school,
and you will never have to worry about whether your advisor has funding or not.
Details about graduate fellowships will be discussed in Section 4.
